---
title: "Spatial Heterogeneity"
author: "Elsi Pöllänen"
date: "2026-06-09"
output: pdf_document
---


## Linear mixed effects model and estimated marginal means

The below described workflow was separately ran for pretreatment and interval debulking surgery sample subsets. 

```{r pretreatment model, echo=FALSE}

library(tidyr)
library(dplyr)
library(lme4)
library(stringr)
library(emmeans)
library(ggplot2)
library(ggcompare)
library(ggbeeswarm)
library(MuMIn)

data_subset <- read.table("~/data.csv")
biomarkers <- c("TACSTD2", "F3", "NECTIN4", "ERBB2", "EGFR", "FOLR1", "CD22")

#------------------------------------------------------------------------------ Linear mixed effects model for each ADC-target:

model_results <- lapply(biomarkers, function(y) {
  formula <- as.formula(paste(y, "~ site + (1|patient) + (1|site:patient)"))
  lmer(formula, data = data_subset)
})

names(model_results) <- biomarkers

#------------------------------------------------------------------------------ Estimated marginal means ADC-target:

emmeans_data <- lapply(model_results, function(model) {
  
  predicted_fixed <- emmeans(model, ~ site)  
  
  emmeans_result <- as.data.frame(predicted_fixed)
  emmeans_result$Model <- deparse(formula(model)) 
  return(emmeans_result)
  
})


# Combining all into one data frame
emmeans_df <- do.call(rbind, emmeans_data)
emmeans_df <- emmeans_df[complete.cases(emmeans_df$emmean),]



emmeans_df$Gene <- sub(" ~.*", "", emmeans_df$Model)

#------------------------------------------------------------------------------ Statististical significance:

significance_p <- lapply(model_results, function(model) {
  
  emms <- emmeans(model, ~ site)  
  
  p_vals <- contrast(emms, method = "pairwise") %>%as.data.frame() 
  
  p_vals$Model <- deparse(formula(model)) 
  return(p_vals)
})

final_p_vals_p <- do.call(rbind, significance_p)

final_p_vals_p$p_adj <- p.adjust(final_p_vals_p$p.value, method = "BH")

significant_pvals_p <- final_p_vals_p[which(final_p_vals_p$p_adj < 0.05),]

significant_pvals_p$Gene <- sub(" ~.*", "", significant_pvals_p$Model)

significant_pvals_p <- significant_pvals_p %>%
  mutate(label = paste0("p = ", signif(p_adj, 3)))

significant_pvals_p <-  emmeans_df %>%
  group_by(Gene) %>%
  summarise(ymax = max(emmean)) %>%
  left_join(significant_pvals_p, by = "Gene") %>%
  mutate(y.position = ymax + 1) %>% as.data.frame()


significant_pvals_p <- significant_pvals_p[complete.cases(significant_pvals_p$label),]

significant_pvals_p$ymin <- str_match(significant_pvals_p$contrast, "([A-Z][a-z]+)[ ][-][ ]([A-Z][a-z]+)")[,2]
significant_pvals_p$ymax <- str_match(significant_pvals_p$contrast, "([A-Z][a-z]+)[ ][-][ ]([A-Z][a-z]+)")[,3]

significant_pvals_p <- significant_pvals_p[complete.cases(significant_pvals_p$contrast),]

significant_pvals_p_splitted <- split(significant_pvals_p, significant_pvals_p$Gene)

results <- lapply(significant_pvals_p_splitted,  function(model) {
 
   a = seq(0, 1, 0.1)
  
    for (i in 1:nrow(model)) {
        model$y.position2[i] <- model$y.position[i] + a[i]
        
    }
  
  return(model)
})

final_results <- do.call(rbind, results)

```

## Sources of heterogeneity

The below described workflow was separately ran for pretreatment and interval debulking surgery sample subsets. 

```{r, Heterogeneity proportions}
#------------------------------------------------------------------------------ Computing tumorsite heterogeneity sources.
#------------------------------------------------------------------------------ Pretreatment.

p_models <- model_results

p_var <- lapply(p_models, function(model){
  
  fixed <- data.frame(fixed = var(predict(model, re.form = NA)))
  random <- as.data.frame(VarCorr(model),comp="Variance")
  
  fixed[, "inter_patient"] <- random$vcov[which(random$grp == "patient")]
  fixed[, "site_patient"] <- random$vcov[which(random$grp == "site:patient")]
  fixed[, "intra_patient"] <- fixed$fixed + fixed$site_patient
  fixed[, "residual"] <- random$vcov[which(random$grp == "Residual")]
  
  return(as.data.frame(fixed))
  
})

pretreatment <- do.call(rbind, p_var)

#------------------------------------------------------------------------------ Saving the total variances for each variable.

pretreatment_var <- pretreatment
pretreatment_var$Gene <- rownames(pretreatment_var)
pretreatment_var <- pretreatment_var[,-which(colnames(pretreatment_var) %in% c("fixed", "site_patient"))]

d_var_long <- pivot_longer(pretreatment_var, cols = -Gene, values_to = "Variance", names_to = "Source of heterogeneity") %>% as.data.frame()
d_var_long$group <- "pretreatment"

#------------------------------------------------------------------------------ Counting the proportions of variance explained by each variable.

pretreatment_proportions <- pretreatment/rowSums(pretreatment[,which(colnames(pretreatment) %in% c( "inter_patient", "intra_patient", "residual"))])
pretreatment_proportions <- pretreatment_proportions[,which( colnames(pretreatment_proportions) %in% c("inter_patient", "intra_patient", "residual"))]

#------------------------------------------------------------------------------ R2 extraction: 

Rsquared <- lapply(p_models, function(model) {
  data.frame(
    r.squaredGLMM(model)
  )
})

Rsquared_results_p <- do.call(rbind, Rsquared) 


Rsquared_results_p <- cbind(Rsquared_results_p, pretreatment_proportions)

colnames(Rsquared_results_p) <- c("R2m_pretreatment", "R2c_pretreatment", "Inter-patient_pretreatment", "Intra-patient_pretreatment", "Residual_pretreatment")

Rsquared_results_p <- round(Rsquared_results_p, digits = 3)

pretreatment_proportions$gene <- rownames(pretreatment_proportions)
pretreatment_proportions$group <- "Pretreatment"

long_pretreatment <- pivot_longer(pretreatment_proportions, cols = -c(gene, group), values_to = "Proportion of variance explained", names_to = "Source of heterogeneity") %>% as.data.frame()

```

## Decision boundary and mean difference of low and high expression regions

```{r folr1 scatterplot of adnex metastasis dynamic,  fig.width=12, fig.height=10, echo=FALSE}

#------------------------------------------------------------------------------ Extracting the patients with both, adnex and metastasis samples: 


x <- as.data.frame(table(data_subset$patient))

for (i in 1:nrow(data_subset)) {
  
  data_subset$nro_samples[i] <- x$Freq[which(x$Var1 == data_subset$patient[i])]
  
}

empty <- data.frame(pat = unique(data_subset$patient))

for (j in 1:length(unique(data_subset$patient))) {
  
  y <- data_subset[which(data_subset$patient == unique(data_subset$patient)[j]),]
  empty[which(empty$pat == unique(data_subset$patient)[j]), "select"] <- ifelse(length(unique(y$site)) > 1, 1,0)
  
}

data_subset <- data_subset[which(data_subset$patient %in% empty$pat[which(empty$select == 1)]),]

longer <- pivot_longer(data_subset, cols = -c("patient", "tissue", "phase", "site", "sample", "nro_samples"), values_to = "Expression", names_to = "Gene") %>% as.data.frame()


#------------------------------------------------------------------------------ Exhaustively matching all available adnex samples with all available metatasis samples for each patient.

splitted_longer <- split(longer,longer$patient)

listed <- lapply(splitted_longer,  function(x) {
  
  all_combos <- expand.grid(Adnex_sample = unique(x$sample[which(x$site == "Adnex")]), Metastasis_sample = unique(x$sample[which(x$site == "Metastasis")]))
  
})

final <- as.data.frame(do.call(rbind, listed))


#------------------------------------------------------------------------------ Adding gene expressions to table and assigning each row to either low or high expression group. This part was run for each target. Only FOLR1 example shown here.


for (i in 1:nrow(final)) {
  
  final$adnex_expression_FOLR1[i] <- longer$Expression[which(longer$sample == final$Adnex_sample[i] & longer$Gene == "FOLR1")]
  final$metastasis_expression_FOLR1[i] <- longer$Expression[which(longer$sample == final$Metastasis_sample[i] & longer$Gene == "FOLR1")]
  
  
}


final$FOLR1_group <- ifelse(
  final$metastasis_expression_FOLR1 > (-final$adnex_expression_FOLR1 + emmeans_df$emmean[which(emmeans_df$Gene == "FOLR1" & emmeans_df$site == "Pretreatment adnex")]*2),
  "above",
  ifelse(
    final$metastasis_expression_FOLR1 < (-final$adnex_expression_FOLR1 + emmeans_df$emmean[which(emmeans_df$Gene == "FOLR1" & emmeans_df$site == "Pretreatment adnex")]*2),
    "below",
    "on_line"
  )
) 


#------------------------------------------------------------------------------ Counting the mean difference for the low and high expression groups for each target.


mean_difference <- as.data.frame(matrix(NA, 7, 2))
colnames(mean_difference) <- c("below", "above")

for (i in 1:length(biomarkers)) {
  
  subset <- final[,grep(biomarkers[i], colnames(final))]
  
  subset$difference <- abs(subset[,grep("adnex", colnames(subset))] - subset[,grep("metastasis", colnames(subset))])
  
  above <- subset[which(subset[,grep("group", colnames(subset))] == "above"),]
  below <- subset[which(subset[,grep("group", colnames(subset))] == "below"),]
  
  rownames(mean_difference)[i] <- biomarkers[i]
  mean_difference$above[i] <- mean(above$difference)
  mean_difference$below[i] <- mean(below$difference)
                  
}


```



